### 2025-01-22 paper/icml25/first/sects/abstract.tex
minors in the flow

\begin{abstract}
Reinforcement learning algorithms have demonstrated success in domains such as robotics and multi-agent systems, 
yet they often exhibit limited sample efficiency in complex environments, 
and a unified approach that effectively addresses both single- and multi-agent settings remains elusive.
To challenge this gap, we introduce a novel framework, \fname{}, inspired by SHAC, a reinforcement learning algorithm designed for differentiable simulators. Unlike SHAC, \fname{} eliminates the requirement for differentiable transition and reward functions by utilizing plain neural networks to estimate the gradients that would otherwise be derived directly from the simulator. These neural networks are trained alongside the policy and value networks of SHAC. We evaluate \fname{} on a set of challenging multi-agent scenarios from VMAS and compare its performance against SHAC and PPO---a current state-of-the-art algorithm for non-differentiable environments. Our results demonstrate that \fname{} significantly outperforms PPO and achieves comparable performance to SHAC, despite not having access to environment gradients.
\end{abstract}

>>>>

\begin{abstract}
Reinforcement learning algorithms have demonstrated success in domains such as robotics and multi-agent systems, 
yet they often exhibit limited sample efficiency in complex environments, 
and a unified approach that effectively addresses both single- and multi-agent settings remains elusive.
To challenge this gap, we introduce a novel framework, \fname{}, inspired by SHAC, which is a reinforcement learning algorithm designed for differentiable simulators. Unlike SHAC, \fname{} eliminates the requirement for differentiable transition and reward functions by utilizing plain neural networks to estimate the gradients that would otherwise be derived directly from simulators. These neural networks are trained alongside the policy and value networks of SHAC. We evaluate \fname{} on a set of challenging multi-agent scenarios from VMAS and compare its performance against SHAC and PPO---a current state-of-the-art algorithm for non-differentiable environments. Our results demonstrate that \fname{} significantly outperforms PPO and achieves comparable performance to SHAC, despite not having access to simulator gradients.
\end{abstract}

### 2025-01-22 paper/icml25/first/sects/experiments.tex
minors + appendix (WIP)
minors in the flow

\paragraphTransport}
The third row of \Cref{fig:experiments} presents a differentiable scenario, transport. In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance. \fname{} converges faster, but SHAC is more consistent due to direct gradient stability. In the second and third columns (3 and 5 agents), PPO fails to converge, while SHAC and \fname{} both complete the task around 10,000 episodes. With more agents, convergence is faster because the task becomes easier (as more agents can push the package). The emergent of cooperation can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package. This is a sign of generalization, as the agents learn to avoid collisions and push the package together.

>>>>

\paragraph{Transport}
The third row of \Cref{fig:experiments} presents a differentiable scenario, transport. In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance. \fname{} converges faster, but SHAC is more consistent due to direct gradient stability. In the second and third columns (3 and 5 agents), PPO fails as in the single-agent case, while SHAC and \fname{} both complete the task around 10,000 episodes. With more agents, convergence is faster because the task becomes easier (as more agents can push the package). The emergent of cooperation can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package. This is a sign of generalization, as the agents learn to avoid collisions and push the package together.

### 2025-01-22 paper/icml25/first/sects/introduction.tex
minors + last related para
prefer Cref on ref
minors in the flow

Learning control policies has important applications in robotics~\cite{Singh22}, autonomous driving~\cite{Elallid22}, swarm intelligence~\cite{Tang21} and many other fields. These scenarios present a variety of different challenges, such as sparse rewards, complex dynamics, and multi-agent interactions. Current approaches often focus on either single- or multi-agent scenarios, lacking a unified solution that works effectively in both settings. Moreover, reinforcement learning algorithms tend to be \emph{sample inefficient}, namely requiring a large number of interactions with the environment to learn a good policy.
\texttt{\color{DarkOrchid}{<- FB: non mi è super chiara questa frase}}. A recently introduced method for the latter problem is SHAC~\cite{Xu22} which exploits differentiable simulators to provide precise gradient information, thereby expediting policy optimization and improving learning efficiency.

Despite its effectiveness, SHAC has two main limitations. First, it requires both a differentiable simulator and a differentiable reward function to operate correctly. Obtaining these can be impractical, particularly in scenarios with inherently sparse rewards. Second, even when a differentiable simulator is available, SHAC has been found to be sensitive to complex dynamics, such as object collisions~\cite{Georgiev24} or interactions in multi-agent systems.

@@

We evaluate our framework, referred to as \fname{}, against two baseline algorithms: PPO~\cite{Schulman17}, a state-of-the-art approach for non-differentiable environments (also widely used in multi-agent scenarios through the MAPPO variant~\cite{DBLP:conf/nips/YuVVGWBW22}), and SHAC~\cite{Xu22}, which is tailored for differentiable environments. The evaluation is conducted using VMAS, a state-of-the-art simulator designed for complex, physics-based multi-agent scenarios~\cite{DBLP:conf/dars/BettiniKBP22}. We compare these algorithms across cooperative tasks involving varying numbers of agents, providing a systematic analysis of how their performance scales with the increasing size of the search space.

>>>>

Learning control policies has important applications in robotics~\cite{Singh22}, autonomous driving~\cite{Elallid22}, swarm intelligence~\cite{Tang21} and many other fields. These scenarios present a variety of different challenges, such as sparse rewards, complex dynamics, and multi-agent interactions. Current approaches often focus on either single- or multi-agent scenarios, lacking a unified solution that works effectively in both settings. Moreover, reinforcement learning algorithms tend to be \emph{sample inefficient}, namely requiring a large number of interactions with the environment to learn a good policy.
%\texttt{\color{DarkOrchid}{<- FB: non mi è super chiara questa frase}}. 
A recently introduced method for the latter problem is SHAC~\cite{Xu22} which exploits differentiable simulators to provide precise gradient information, thereby expediting policy optimization and improving learning efficiency.

Despite its effectiveness, SHAC has two main limitations. First, it requires both a differentiable simulator and a differentiable reward function to operate correctly. Obtaining these can be impractical, particularly in scenarios with inherently sparse rewards. Second, even when a differentiable simulator is available, SHAC has been found to be sensitive to complex dynamics, such as object collisions~\cite{Georgiev24} or interactions in multi-agent systems. Backpropagation through these dynamics has been found, in general, to lead to unstable gradients~\cite{Bengio94,Metz21}.

@@

We evaluate our framework, referred to as \fname{}, against two baseline algorithms: PPO~\cite{Schulman17}, a state-of-the-art approach for non-differentiable environments (also widely used in multi-agent scenarios through the MAPPO variant~\cite{DBLP:conf/nips/YuVVGWBW22}), and SHAC~\cite{Xu22}, which is tailored for differentiable simulators. The evaluation is conducted using VMAS, a state-of-the-art simulator designed for complex, physics-based multi-agent scenarios~\cite{DBLP:conf/dars/BettiniKBP22}. We compare these algorithms across cooperative tasks involving varying numbers of agents, providing a systematic analysis of how their performance scales with the increasing size of the search space.
